### 2014-08-19 drafts/ApproximatePCFG/ApproximatePCFG.tex
Implemented Boundary Probabilties but is way too slow. Need to optimize with HashTable.

\usepackage[top=2in, bottom=1.5in, left=1in, right=1in]{geometry}
\usepackage{amsmath}

@@

@@

\begin{algorithm}
\caption{The Approximate Outside Algorithm.\label{alg:approxOutside}}
\begin{algorithmic}[2]
\Procedure {$ApproxOutside$}{$s$, $G$, $S$}
  \State $n \leftarrow length(s)$
  \State $\beta[S, 1, n] \leftarrow 1$ \Comment initialize $\beta$ recursion, unless $n==1$ in which case we are done
  \State $open \leftarrow [(S, 1, n)]$ \Comment initialize list that will hold nodes to be processed
  \While {$open \neq NIL$}
  \State $children \leftarrow []$ \Comment initialize list that holds next level of nodes to be processed
   \For {$(A, i, j) \in open$ \Comment $i \neq j$}
    \For {$A \leftarrow B\, C \in rules(G)$} \Comment iterate over binary rules headed by $A$
      \For {$k \leftarrow i, \dots, j - 1$}
        \If {$\alpha[B, i, k] \neq 0$}
          \State if $\beta[C, k+1, j] = NIL$, set to $0$
          \State $\beta[C, k+1, j] \leftarrow \beta[C, k+1, j] + \psi(A \rightarrow B C) \alpha[B, i, k] \beta[A, i, j]$
          \State add $(C, k+1, j)$ to $children$
        \EndIf

@@

\begin{algorithm}
\caption{The Approximate Outside Algorithm.\label{alg:approxOutside}}
\begin{algorithmic}[2]
\Procedure {$ApproxOutside$}{$s$, $G$, $S$}
  \State $OPEN \leftarrow []$ \Comment initialize OPEN FIFO queue
  \State $n \leftarrow length(s)$
  \State $\beta(S, 1, n) \leftarrow 1$
  \State \Call{AddChildrenToOpen}{$S, 1, n, \beta(S, 1, n)$}
  \While {$OPEN$ is not empty}
    \State $node \leftarrow pop(OPEN)$ where $rule_{node} = (A \rightarrow B C, i, k, j) = fst(node)$ and $pa_\beta= snd(node)$
    \State $a_\ell \leftarrow \alpha[B, i, k]$
    \State $a_r \leftarrow \alpha[C, k+1, j]$
    \If {$a_r \neq NIL$}
      \If {$\beta[B, i, k] = NIL$}
         $\beta[B, i, k] \leftarrow 0$
      \EndIf
      \State $\beta[B, i, k] \leftarrow \beta[B, i, k] + pa_\beta a_\ell \psi(A \rightarrow B C)$ \Comment compute beta value and add to current beta value 
      \State \Call{AddChildrenToOpen}{$B, i, k, \beta(B, i, k)$}
    \EndIf
    \If {$a_l \neq NIL$}
      \If {$\beta[C, k+1, j] = NIL$}
         $\beta[C, k+1, j] \leftarrow 0$
      \EndIf
      \State $\beta[C, k+1, j] \leftarrow \beta[C, k+1, j] + pa_\beta a_r \psi(A \rightarrow B C)$ \Comment compute beta value and add to current beta value 
      \State \Call{AddChildrenToOpen}{$C, k+1, j, \beta(C, k+1, j)$}
    \EndIf
  \EndWhile  
\EndProcedure

\Statex
\Procedure {AddChildrenToOPEN}{A, i, j, betaVal}
  \State $\{c_i\}\leftarrow \Call{GetChildren}{A, i, j}$ \Comment Get AND-node children
  \State insert pairs $\{(c_i, betaVal\}_i)$ into OPEN
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\section{A simple upper bound}
We will present a simple upper bound on a slightly restricted set of
PCFGs which is based on the following intuition: if I am a nonterminal
$A$ and I know that I need to yield a string $x_1 \dots x_n$, then
this is going to be at least as hard as starting at $A$ and generating
any one of the elements in $x_1 \dost x_n$. That is, the highest
probability single word in my yield will have a higher probability of
being produced, then any longer span.

Suppose our PCFG has potentials less than or equal to one. This is the
case in the commons situation in which the PCFG is normalized, or if,
as in variational approximations to the Hierarchical Diirichlet PCFG,
our PCFG is unnormalized with values summing to less than one. Now we
define a function $g(A, w, m)$ for nonterminal $A$, terminal $w$ and
integer $w>0$, which we claim is greater the weight our PCFG would
assign to any parse tree rooted at $A$ and yielding any span of length
$m$ containing $w$. Specifically, consider any such parse tree, and
set all weights not on the path between $A$ and $w$ to 1. The total
weight assigned to the parse tree is the product of the weights on its
edges, so the resulting parse tree will have a strictly greater score.

\begin{algorithm}
\caption{An O(n^2}
\begin{algorithmic}[2]
\Procedure {$GO_{OR}$}{$A$, $i$, $j$}

\EndProcedure
\end{algorithmic}
\end{algorithm}

>>>>

\usepackage[top=2in, bottom=1.5in, left=1in, right=1in]{geometry}
\usepackage{amssymb,amsmath}

@@

\usepackage{natbib}

\makeatletter
\newcommand{\xRightarrow}[2][]{\ext@arrow 0359\Rightarrowfill@{#1}{#2}}
\makeatother

\newcommand{\derives}{\xRightarrow{*}}

@@

\begin{algorithm}
\caption{The Approximate Outside Algorithm.\label{alg:approxOutside}}
\begin{algorithmic}[2]
\Procedure {$ApproxOutside$}{$s$, $G$, $S$}
  \State $n \leftarrow length(s)$
  \State $\beta[S, 1, n] \leftarrow 1$ \Comment initialize $\beta$ recursion, unless $n==1$ in which case we are done
  \For {$\ell \gets 1, \dots, n-1$}
    \For {$i \gets \ell$}
      \State $j \leftarrow i + \ell$
    \EndFor
  \EndFor

  \State $open \leftarrow [(S, 1, n)]$ \Comment initialize list that will hold nodes to be processed
  \While {$open \neq NIL$}
  \State $children \leftarrow []$ \Comment initialize list that holds next level of nodes to be processed
   \For {$(A, i, j) \in open$ \Comment $i \neq j$}
    \For {$A \leftarrow B\, C \in rules(G)$} \Comment iterate over binary rules headed by $A$
      \For {$k \leftarrow i, \dots, j - 1$}
        \If {$\alpha[B, i, k] \neq 0$}
          \State if $\beta[C, k+1, j] = NIL$, set to $0$
          \State $\beta[C, k+1, j] \leftarrow \beta[C, k+1, j] + \psi(A \rightarrow B C) \alpha[B, i, k] \beta[A, i, j]$
          \State add $(C, k+1, j)$ to $children$
        \EndIf

@@

@@

The algorithm presented provides two degrees of freedom to the user: the first is a choice of upperbound function $U$, and the second is a way to order the children of a node. The tighter the upperbound, the better an approximation we get. The better the ordering of the children, the less often we need to explore derivations of low relative weight.

\section{Ordering heuristics}
How we order the children of a node is critical to the speed-up this
aproximate algorithm gets. After all, if there is a child whose
derivation dominates all others and we accidentally put it last, then
we will not be able to short-circuit at all; if, instead, we put it
first, then we will likely only have to explore that child.

One can image various choices of ordering functions, with perhaps the
simplest being to pick the rules in order of decreasing weight. This,
however, only gives us an ordering on the rules, not on where we
should split the span of the child. One reasonable suggestion,
especially for right branching structures like English parse trees, is
to start with the smallest left child spans first and increase from
there.

These heuristics, however, are very weak, and do not take into
consideration the content span at all. In the following we propose a
simple heuristic with child ordering that takes into account the
bottom-up probability of the content but whose complexity is constant
in the size of the span. The basic intutiion behind this heuristic is
that if one wants to evaluate the probability that, e.g., a noun
phrase will yield that fragmant ``saw the boy with,'' one can gain a
lot of information by calculating the probability that the phrase will
start and end with ``saw'' and ``with,'' respectively.

Some notation: let's denote by $\gamma(A, x)$ the probability that a
nonterminal $A$ derives any string with prefix $x$. Likewise, let's
denote by $\theta(A, x)$ the proabability that a nonterminal $A$
derives any string with suffix $x$. We will write this using the
following notation:

\begin{align}
  \gamma(A, x) &\triangleq p(A \xRightarrow{*} x\dots) \label{eq:gammaDef}\\
  \theta(A, x) &\triangleq p(A \xRightarrow{*} \dots x)  \label{eq:thetaDef}
\end{align}

$\gamma(A, x)$ is just the prefix probability defined for a single
lexical item. This can be defined as:
\begin{align}
  p(A \derives x\dots) &= \sum_{A \rightarrow BC} p(B \derives x \dots) p(A \rightarrow B C) 
    + p(A \rightarrow x)\\
  p(A \derives \dots x) &= \sum_{A \rightarrow CB} p(B \derives \dots x) p(A \rightarrow C B) 
    + p(A \rightarrow x)
\end{align}

If we define the matrices $\Gamma$ and $\Theta$ where $\Gamma_{A, B} = \gamma(A, B)$
and $\Theta_{A, B} = \theta(A, B)$. then we can rewrite these recursions

\begin{align}
  \Gamma &= \Phi^\ell \Gamma + \Psi \\
  \Theta &= \Phi^r \Gamma  + \Psi \\
  \text{where }\Phi^\ell_{A, B} &\triangleq \sum_{A \rightarrow B C} p(A \rightarrow B C)\\
               \Phi^r_{A, B} &\triangleq \sum_{A \rightarrow C B} p(A \rightarrow C B)\\
               \Psi_{A, x} &\triangleq p(A \rightarrow x).
\end{align}

This means that we can calculate $\Gamma$ and $\Phi$ by solving the linear systems

\begin{align}
  (1-\Phi_\ell) \Gamma &= \Psi\\
  (1-\Phi_r) \Theta &= \Psi.\\
\end{align}

The complexity of these calculations is $K^3$, where $K$ is the number
of symbols in the grammar.

From $\Gamma$ and $\Theta$ we can compute the boundary probability: 
\begin{align}
  p(A \derives x \dots y) &= p(A \rightarrow x) \mathbf{1}(x = y) 
     + \sum_{A \rightarrow B C} p(A \rightarrow B C) p(B \derives x \dots) p(C \derives \dots y)\\
  p(A \derives x \dots y) &= p(A \rightarrow x) \mathbf{1}(x = y) 
     + \sum_{A \rightarrow B C} p(A \rightarrow B C) \Gamma_{B, x} \Theta_{y, C}
\end{align}

Here, as is often the case in computations that make use of heuristic
scores, it may be advantageous to precompute all the heuristic values
so that lookup is $O(1)$.

\section{Appendix}
\subsection{Existence of $\Gamma$ and $\Theta$}
\textbf{Claim: } The matrices $\Gamma$ and $\Theta$ defined by
Equations~\ref{eq:gammaDef}-\ref{eq:thetaDef} exist.  \textbf{Proof: }
The $\Gamma$ and $\Theta$ matrices will not exist only if
$1-\Phi^\ell$ and $1-\Phi^r$ are singular. To show that they are not
singular we will show that there is no non-zero linear combination of
their columns that sums to zero. The proofs are nearly identical for
$1-\Phi^\ell$ and $1-Phi^r$.

Suppose, by contradiction, that their is some linear combination $a_1,
..., a_K$ such that $(1-\Phi^\ell) a = 0$. This implies that each $a_i
= \sum_{j} \Phi^\ell_{i, j} a_j$. Recall that the element of
$\Phi^\ell_{A, B} = \sum_{A \rightarrow B C} \p(A \rightarrow B C)$,
where $A$ and $B$ are nonterminals. For the PCFG defined by these
probabilities to be valid, $\Phi^\ell_{A,B}$ must be less than
one. Otherwise, there would be a nonzero probability of an infinite
string. Taking the absolute values of both sides of the equation
above, we see that $|{a_i}| \leq \sum_{j} \Phi^\ell_{i, j} |
a_j|$. But if this is true, then  TODO

\bibliographystyle{bibliographystyle}
\bibliography{main}
